Pangenome graph layout by Path-Guided Stochastic Gradient Descent

### 2023-07-17 manuscript/main.tex

    \subsection{Performance}
	We apply PG-SGD to the recent draft human pangenome reference \citep{Liao2023} to show the scalability and practicality of the algorithm with a real pangenome.
	Experiments were conducted on a cluster with 24 \texttt{Regular} nodes (32 cores / 64 threads AMD EPYC 7343 processors with 512 GB RAM each) and 4 \texttt{HighMem} nodes (64 cores / 128 threads AMD EPYC 7513 processors with 2048 GB RAM each).
	We downloaded pangenome graphs for each autosome (24 in total) and for the mitochondrial DNA.
	Each graph represents 90 whole human haplotypes: 44 diploid individuals plus the GRCh38 \citep{Schneider2017} and CHM13 \citep{Nurk_2021} haploid human references.
	See Supplementary Table~\ref{tab:layout} for graph statistics.
	When applied to these pangenome graphs using one \texttt{Regular} node for each calculation, PG-SGD obtains the graph layouts in \FIXME{XXXX minutes/hours on average}, with the slowest chromosome being chromosome 16 (Supplementary Table~\ref{tab:layout}).
	This is expected since chromosome 16 presents one of the highest levels of segmentally duplicated sequence among the human autosomes \citep{Martin2004}.
	Repetitive sequences lead to graph nodes with a very high number of path traversals, which are computationally expensive to work with \citep{Guarracino2022}.
	Memory consumption is \FIXME{XXXX GB of RAM on average}, with the memory peak occurring with chromosome 1, the biggest one in humans.
	Given its scalability, we even applied PG-SGD to the full graph with all chromosomes together using a \texttt{HighMem} node (Supplementary Table~\ref{tab:layout}).
	Overall, PG-SGD is ~\FIXME{XX}X faster than BandageNG \FIXME{HOW TO CITE IT?}, the current state-of-the-art for graph visualization, while using ~{}X less memory.

@@

	%We are grateful to members of the HGSVC and HPRC production teams for their development of resources used in our exposition.
	%We thank the authors of the pangenome resources made available on GenBank which have made our experiments possible.
	A special thanks goes to Matthias Seybold from the Quantitative Biology Center for maintaining the Core Facility Cluster.

@@

	S.H. acknowledges funding from the Central Innovation Programme (ZIM) for SMEs of the Federal Ministry for Economic Affairs and Energy of Germany.
	S.N. acknowledges Germany’s Excellence Strategy (CMFI), EXC-2124 and (iFIT)—EXC 2180–390900677.
	This work was supported by the BMBF-funded de.NBI Cloud within the German Network for Bioinformatics Infrastructure (de.NBI) (031A532B, 031A533A, 031A533B, 031A534A, 031A535A, 031A537A, 031A537B, 031A537C, 031A537D and 031A538A).
	A.G. acknowledges efforts by Nicole Soranzo to establish a pangenome research unit at the Human Technopole in Milan, Italy.
	JNM.S., J.L., and Z.Z. acknowledge funding from the NSF PPoSS Award \#2118709.

>>>>

    \subsection{Performance}
	We apply PG-SGD to the recent draft human pangenome reference \citep{Liao2023} from the Human Pangenome Reference Consortium (HPRC) to show the scalability of the algorithm with a real pangenome.
	Experiments were conducted on a cluster with 24 \texttt{Regular} nodes (32 cores / 64 threads with two AMD EPYC 7343 processors with 512 GB RAM each) and 4 \texttt{HighMem} nodes (64 cores / 128 threads with two AMD EPYC 7513 processors with 2048 GB RAM each).
	We downloaded pangenome graphs for each autosome (24 in total) and for the mitochondrial DNA.
	Each graph represents 90 whole human haplotypes: 44 diploid individuals plus the GRCh38 \citep{Schneider2017} and CHM13 \citep{Nurk_2021} haploid human references.
	See Supplementary Table~\ref{tab:layout} for graph statistics.
	When applied to these pangenome graphs using one \texttt{Regular} node for each calculation, PG-SGD obtains the graph layouts in \FIXME{XXXX minutes/hours on average}, with the slowest chromosome being chromosome 16 (Supplementary Table~\ref{tab:layout}).
	This is expected since chromosome 16 presents one of the highest levels of segmentally duplicated sequence among the human autosomes \citep{Martin2004}.
	Repetitive sequences lead to graph nodes with a very high number of path traversals, which are computationally expensive to work with \citep{Guarracino2022}.
	Memory consumption is \FIXME{XXXX GB of RAM on average}, with the memory peak occurring with chromosome 1, the biggest one in humans.
	Given its scalability, we even applied PG-SGD to the full graph with all chromosomes together using a \texttt{HighMem} node (Supplementary Table~\ref{tab:layout}).
	Overall, PG-SGD is ~\FIXME{XX}X faster than BandageNG \FIXME{HOW TO CITE IT?}, the current state-of-the-art for graph visualization, while using ~{}X less memory.

@@

	%We are grateful to members of the HGSVC and HPRC production teams for their development of resources used in our exposition.
	%We thank the authors of the pangenome resources made available on GenBank which have made our experiments possible.
	We thank Matthias Seybold from the Quantitative Biology Center for maintaining the Core Facility Cluster.

@@

	S.H. acknowledges funding from the Central Innovation Programme (ZIM) for SMEs of the Federal Ministry for Economic Affairs and Energy of Germany.
	S.N. acknowledges Germany’s Excellence Strategy (CMFI), EXC-2124 and (iFIT)—EXC 2180–390900677.
	This work was supported by the BMBF-funded de.NBI Cloud within the German Network for Bioinformatics Infrastructure (de.NBI) (031A532B, 031A533A, 031A533B, 031A534A, 031A535A, 031A537A, 031A537B, 031A537C, 031A537D and 031A538A).
	A.G. acknowledges efforts by Nicole Soranzo to establish a pangenome research unit at the Human Technopole in Milan, Italy.
	JNM.S., J.L., Z.Z., P.P., and E.G. acknowledge funding from the NSF PPoSS Award \#2118709.
